Debugging by \textit{lastChange}

### 2010-11-19 papers/pldi2011/sigplanconf-template.tex

\subfigure[The result of \textit{lastChange} query for
  \texttt{myCondition.value}. To evaluate an expression (e.g., oldValue) at this point, developer can
  enter the expression in the watch box and after re-execution the result is available.
  The opened list on the top of the left panel shows the visited execution points. Clicking on each point in
  the list shows the corresponding code and
  data.]{\label{fig:example4}\includegraphics[width=1.0\textwidth%,height=.25\textheight
    ]{4-lastChange2.jpg}}

@@

\subsection{\texttt{objectId()} operation}
\texttt{objectId(obj)} first checks the argument \texttt{obj} for a property \texttt{\_objectId}.
It returns the value if this property is already defined,
otherwise it generates a new id and sets this property. The value of the id is simply an integer incremented 
for each new \texttt{\_objectId} needed. To assign the id to the object we 
set the id as a regular object property. This might change
the program behaviour due to the appearance of this property in 
\texttt{for(property in object)} loops. \texttt{defineProperty} is a standard JavaScript
function which is available in the next version of Firefox\footnote[1]{https://developer.mozilla.org/en/JavaScript/Reference/Global\_Objects\\/Object/defineProperty}. \texttt{defineProperty} receives 
an argument which specifies whether the property is enumerable or not. By setting the enumerable
field \texttt{false} for \texttt{\_objectId}, this property will not
apear in \texttt{for(property in object)} loops and therfore has no
effect on the program execution. Note that \texttt{\_objectId} is
not set for all objects but only those objects that need an id.

@@

Throughout this section we implicitly assumed that the object will be created at 
the same location in the next execution. So, what does happens if it
is not true? Well, once the execution reaches the reproduction point, it
reveals that the object has been created in different location. This time,
debugger re-executes the program considering both locations as possible 
object creation locations.

@@

\section{Reproducible Non-deterministic Execution}

By construction \textit{lastChange} works whenever conventional
breakpoint debugging works. By that we mean that, since we use
conventional breakpoint debugging technology under the covers,
\textit{lastChange} works whenever breakpoint debugging
works. However, there are cases where breakpoint debugging fails but
\textit{lastChange} works.  These are cases of \textit{reproducible
  non-deterministic} execution.

A bug is \textit{reproducible} for a developer when the developer can
start from a determined initial state, operate on the program with a
list of actions, and reproduce the symptoms of the bug. The details of
the execution can change each time we re-execute the buggy program,
but the buggy result is the same.  The reproducibility of the bug
means that the defect is very unlikely to depend on the order of
events during the execution. When the developer selects a value and
asks for \textit{lastChange}, the developer is expressing the hypothesis that
the value is related to the defect. Therefore the reproducibility of
the bug makes it very unlikely that the \textit{lastChange} will
depend upon order of events during execution.

Since \textit{lastChange} does not rely on determinism, some bugs that
confound breakpoint debuggers will be easier to find with
\textit{lastChange}.  Consider the example in 
Figure ~\ref{fig:counter-example}. It shows a function which processes an
array. The array contains numbers except one item which is
\texttt{undefined} and it causes a bug in line 30. There is a call to
\texttt{randomPermutation} function in line 11 which randomly
permutes the array item. So at every execution the
\texttt{undefined} item will be in a new place. Calling
\textit{lastChange} on \texttt{x} at the place the bug happens, gives a
point which shows line 14. Although this point exists at every
re-execution but it can not be identified by a conditional breakpoint.

@@

Thus in the case of a non-deterministic program, a \textit{lastChange}
is not equivalent to any series of conventional watchpoints or
breakpoints. The extra power of \textit{lastChange} derives from its
analysis of the entire execution up to the reproduction point and its
memory of previous work on the defect.  Each time we re-execute a
non-deterministic program, the details of execution instruction order
may change. For example, if we record the source code lines every time
a conventional watchpoint hits, the record may differ each time we
re-execute. But \textit{lastChange} does not ask the developer to
consider these single points.  Rather we analyze all of the
watchpoints in an entire execution at the reproduction point where we
know the defect has already occurred and only show the final result to
the developer.

Of course there can be cases where the defect is reproducible but the
value from \textit{lastChange} is not. As in conventional breakpoint
debugging, we can only determine this empirically, by observing
multiple executions with different values. Unlike conventional
breakpoint debugging, implementations of \textit{lastChange} readily
help in detecting such cases: as we work back through a series of
\textit{lastChange} requests we can compare values with previous
executions. Different values on different executions will signal that
the execution is not deterministic. In future we hope to compare
queries from successive executions as a tool for learning about
non-deterministic executions.

@@

\subsection{Developers need an operation like \protect\textit{lastChange} }

>>>>

\subfigure[The result of \textit{lastChange} query for
  \texttt{myCondition.value}. To evaluate an expression (e.g., oldValue) at this point, developer can %TODO should we remove it?
  enter the expression in the watch box and after re-execution the result is available.
  The opened list on the top of the left panel shows the visited execution points. Clicking on each point in
  the list shows the corresponding code and
  data.]{\label{fig:example4}\includegraphics[width=1.0\textwidth%,height=.25\textheight
    ]{4-lastChange2.jpg}}

@@

\subsection{\texttt{objectId()} operation}
\texttt{objectId(obj)} first checks the argument \texttt{obj} for a property \texttt{\_objectId}.
It returns the value if this property is already defined,
otherwise it generates a new id and sets this property. The value of the id is simply an integer incremented 
for each new \texttt{\_objectId} needed. To assign the id to the object we 
set the id as a regular object property. This might change
the program behaviour due to the appearance of this property in 
\texttt{for(property in object)} loops. \texttt{defineProperty} is a standard JavaScript
function which is available in the next version of Firefox\footnote[1]{https://developer.mozilla.org/en/JavaScript/Reference/Global\_Objects\\/Object/defineProperty}. \texttt{defineProperty} receives 
an argument which specifies whether the property is enumerable or not. By setting the enumerable
field \texttt{false} for \texttt{\_objectId}, this property will not
apear in \texttt{for(property in object)} loops and therefore has no
effect on the program execution. Note that \texttt{\_objectId} is
not set for all objects but only those objects that need an id.

@@

Throughout this section we implicitly assumed that the object will be created at 
the same location in the next execution. So, what does happen if it
is not true? Well, once the execution reaches the reproduction point, it
reveals that the object has been created in different location. This time,
debugger re-executes the program considering both locations as possible 
object creation locations.

@@

\section{Reproducible Non-deterministic Execution}
We claimed that the only prerequisite for \textit{lastChange} function is bug
reproducibility. A bug is \textit{reproducible} for a developer when the developer can
start from a determined initial state, operate on the program with a
list of actions, and reproduce the symptoms of the bug. The details of
the execution can change each time we re-execute the buggy program,
but the buggy result is the same.  The reproducibility of the bug
means that the defect is very unlikely to depend on the order of
events during the execution. When the developer selects a value and
asks for \textit{lastChange}, the developer is expressing the hypothesis that
the value is related to the defect. Therefore the reproducibility of
the bug makes it very unlikely that the \textit{lastChange} will
depend upon order of events during execution. In this section we discuss issues regarding to non-determinism.

\subsection{Pausing at \textit{lastChange} Result}
A \textit{lastChange} result can be recognized early when it occurs if it can
be defined by a conditional breakpoint. A proposal for the condition is the event index
in the history of change events. This condition obviously works for deterministic 
executions. Therefore, if developer specifies an execution determinitic, every lastChange
will become pausable. Moroever, it works in a non-deterministic execution if non-determinism
has no effect on the event index. If in several executions, the \textit{lastChange} result happens at
the same index in the list of change events, it is very likely that in the next execution, the 
same index be the result. Debugger can propose it as a pausable point to the developer or optimizes data
collection by focusing on this point. The correctness of this guess can be verified at the reproduction
point.

There are cases where finding the appropriate condition requires developer's knowledge.
Consider the example in Figure~\ref{fig:counter-example}. It shows a function which processes an
array. The array contains numbers except one item which is
\texttt{undefined} and it causes a bug in line 30. There is a call to
\texttt{randomPermutation} function in line 11 which randomly
permutes the array item. So at every execution the
\texttt{undefined} item will be in a new place. Calling
\textit{lastChange} on \texttt{x} at the place the bug happens, gives a
point which shows line 14. Although this point exists at every
re-execution but it can not be identified by any condition excepting
the value of item.

@@

\subsection{\textit{lastChange} Result Consistency}
Each time we re-execute a non-deterministic program, 
the details of execution instruction order may change. 
For example, if we record the source code lines every time
a conventional watchpoint hits, the record may differ each time we
re-execute. But \textit{lastChange} does not ask the developer to
consider these single points. Rather it analyzes all of the
change events in an entire execution at the reproduction point where the bug
has already occurred and shows the final result to the developer.

Of course there can be cases where the bug is reproducible but the
value from \textit{lastChange} is not. As in conventional breakpoint
debugging, we can only determine this empirically, by observing
multiple executions with different values. Unlike conventional
breakpoint debugging, implementations of \textit{lastChange} readily
help in detecting such cases: as we work back through a series of
\textit{lastChange} requests we can compare values with previous
executions. Different values on different executions will signal that
the execution is not deterministic. Consider the example in Figure~\ref{fig:nondeterministic-values}. 
Assume that at the first exeuction, developers sees that \texttt{a} is \texttt{true} and asks
for the \textit{lastChange} of \texttt{a} at breakpoint. In the re-exeuction, \texttt{a} is \texttt{false},
and \texttt{b} is \texttt{b} is \texttt{true} at the reproduction point. It means that \texttt{lastChange}
result shows line 10, which is a correct answer for this execution but not for the previous one.
When user asks for the \textit{lastChange} on variable \texttt{a}, debugger stores \texttt{a} value and compares
it to the \texttt{a} value in the next executions. So if this value is different, debuggers inform the developer
that the \textit{lastChange} query is made on a different value. In cases that the value is not a primitive but an object,
the properties of the object examined by the developer are stored and compared.

\begin{figure}[htp]
\begin{verbatim}
10 a = false;
11 b = false;
...
20 if ( random() ) { a = true; } else { b = true; }
...
30 if (a || b) { bug();  /* breakpoint */ }
\end{verbatim}
\caption{\textit{lastChange} on non-deterministic values.}
\label{fig:nondeterministic-values}
\end{figure}

@@

%TODOWe have also developed a prototype implementation of the debugger control part of lastChange for Java. The main differences (one or two sentences). We will report on this implementation in a later publication
\subsection{Developers need an operation like \protect\textit{lastChange} }
